---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```



```{r load_libraries, message=FALSE, warning=FALSE, include=FALSE}

# load libraries

library(readr)
library(dplyr)
library(stringr)
library(magrittr)
library(ggplot2)
library(readr)
library(tidytext)
library(tidyr)
library(tidyverse)
library(ggthemes)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(mapsf)
library(colorspace)
library(shiny)
library(shinydashboard)

```



```{r setup, message=FALSE, warning=FALSE, include=FALSE}
#> read in datasets 

# pooled study branch

# global variables

who_pm2.5_guideline <- 5

# epi studies analysis raw sheet
epi <- readxl::read_xlsx("./data-raw/pm2.5_distribution/AQLI_Epidemiology Literature Research.xlsx", sheet = "AnalysisDatasetPM2.5MortalityAn")

# AQLI color file
aqli_color <- read_csv("./data-raw/pm2.5_distribution/color.csv") 

# global operations
`%notin%` <- Negate(`%in%`)

#> change default columns types
epi$cohort_size <- as.numeric(epi$cohort_size)
epi$study_start_year<- as.numeric(epi$study_start_year)
epi$study_end_year <- as.numeric(epi$study_end_year)
epi$pm2.5_exposure_ll <- as.numeric(epi$pm2.5_exposure_ll)
epi$pm2.5_exposure_ul <- as.numeric(epi$pm2.5_exposure_ul)
epi$mean_pm2.5 <- as.numeric(epi$mean_pm2.5)
epi$sd_pm2.5 <- as.numeric(epi$sd_pm2.5)
epi$cohort_age_ll <- as.numeric(epi$cohort_age_ll)
epi$cohort_age_ul <- as.numeric(epi$cohort_age_ul)

#> add useful columns and filter out some studies (e.g. pooled studies, meta analysis)

epi <- epi %>%
  mutate(study_duration = (study_end_year - study_start_year) + 1)

#> Calculations needed for all sections above the "Results section".

# percent of population not in compliance with the WHO guideline
num_people_above_who <- aqli_color %>%
  filter(pm2020 > who_pm2.5_guideline) %>%
  summarise(tot_pop = sum(population, na.rm = TRUE)) %>%
  unlist()

# world population
world_pop <- aqli_color %>%
  summarise(tot_pop = sum(population, na.rm = TRUE)) %>%
  unlist()
  
percent_people_above_who <- round((num_people_above_who/world_pop)*100, 1)

```


# Epidemiological Studies: Meta Analysis

<!-- badges: start -->
<!-- badges: end -->

This analysis is our best first attempt to capture all epidemiological cohort studies (>`r min(epi$cohort_size, na.rm = TRUE)` people; max cohort size: `r round(max(epi$cohort_size, na.rm = TRUE)/1000000, 1)` million), that are relatively long term ((> `r min(epi$study_duration, na.rm =TRUE)` years, max study duration: `r max(epi$study_duration, na.rm =TRUE)` years) and measure the impact of ambient PM~2.5~, PM~10~, TSP, Ultra Fine Particulate Matter on Mortality (cause-specific, all cause, pre-mature)/Life Expectancy published between `r min(epi$publishing_year)` and present*, findable in available peer-reviewed literature. Hereafter, referred to as **AQ epi studies** for short.


We are seeking to make this analysis as current, complete and error-free as possible and view it as a continual work in progress. We would appreciate the air quality community’s comments, corrections, and suggestions. Please contact **aqli-epic@uchicago.edu** or leave a comment in [this](https://github.com/aqli-epic/epi.meta.analysis) GitHub repository. 

## Purpose

The purpose of this analysis is to understand the landscape of epidemiological research on the relationship between PM~2.5~, PM~10~, TSP, Ultrafine Particulate Matter and Mortality (all types, as specified above)/Life Expectancy and to surface demographic, geographic, or other trends that may exist in the current state of literature. While the overall arc of the relationship between these pollutants and human health is clear to take action, such trends can help the field reflect on itself, take stock of any biases or gaps – and point toward future research and policy opportunities.


## Why do epidemiological studies on air pollution and mortality matter?

While global estimates of air pollution’s toll on public health vary, they all point in the same direction: air pollution poses one of the largest health risks on the planet to humans [[Paper1](https://www.pnas.org/doi/10.1073/pnas.1300018110), [Paper2](https://www.pnas.org/doi/full/10.1073/pnas.1616784114), [Paper3](https://pubs.acs.org/doi/pdf/10.1021/acs.estlett.8b00360), [Paper4](https://pubs.acs.org/doi/pdf/10.1021/acs.estlett.8b00360)]. Epidemiological studies on air pollution and mortality help us understand the burden of air pollution on human health at global, national and regional levels. According to [Vahlsing and Smith (2012)](https://link.springer.com/content/pdf/10.1007/s11869-010-0131-2.pdf), these sorts of studies can also help countries take policy action, pushing forward and shaping national-level ambient air quality standards.

The burden of air pollution across the world is also not uniform. While `r percent_people_above_who` percent of the world population is out of compliance with the latest WHO annual PM~2.5~ guideline of `r who_pm2.5_guideline` µg/m³, there is huge variation in the quality of air one breathes. 


## Inclusion/Exclusion Criteria for studies

  * The underlying [**__analysis dataset__**](https://docs.google.com/spreadsheets/d/1AljEJhNPLWX_8xRbT_HJuERBpbQt_QGixgJ9jEzFyQw/edit#gid=2082201996) for this meta analysis focuses only on those papers that study the link between **PM~2.5~**, **PM~10~**, **TSP**, **Fine Particulate Matter** and **__Mortality/Life Expectancy__**.  In addition to the analysis dataset, there is a [**__master dataset__**](https://docs.google.com/spreadsheets/d/1AljEJhNPLWX_8xRbT_HJuERBpbQt_QGixgJ9jEzFyQw/edit#gid=0) that expands on the analysis dataset to include other useful information. The idea behind the master dataset is to record any additional details (whether additional facts about the paper, or details on other pollutants studied in the paper) that will not be a part of the main data analysis exercise. 
  

  * The analysis dataset excludes the following types of papers: meta-analysis, unpublished papers, papers studying the effects of indoor air pollution on Mortality/Life Expectancy, papers forecasting future air pollution/life expectancy/mortality. But, the master dataset lists all of these studies.

  * In Multi-Country (pooled) studies, one entry (one row in the analysis dataset) is recorded for each country in papers where country level data is available. There are some pooled studies where country level data is not available, but rather data is available for a custom region (e.g. South-East Asia), such pooled studies are excluded from the analysis dataset. But, the master dataset still lists all of these additional studies for reference.
  
  * In cases where the same cohort is studied by different research groups at different point in time/using different methods: we have included all of those studies.
  
  * Papers studying the health effects of pollution segregated by sectors (source apportionment type studies) are excluded from this analysis. 
  
  * All papers studying the measure the impact of pollutants on DALY's, have not been included in this analysis.
  


## Other important points


  * In papers, where the minimum PM~2.5~ concentration was not reported, we have assumed the lowest available percentile data available on PM2.5 concentration as the minimum concentration.
  
  * * In many papers, only one of the mean PM~2.5~ or PM~2.5~ range is reported but not both. In these cases, wherever the data is not available (whether it is mean PM2.5 or PM2.5 range) we have recorded a NA. Apart from this, anywhere we couldn't find data, we have recorded a NA.


  * In multi-pollutant studies where different pollutants have been studied over different periods of time, we have chosen the specific time period that corresponds to the PM~2.5~ pollutant (in cases where we PM~2.5~ is not present, we have recorded the time period corresponding to one of the other pollutants).


  * We have recorded one mean PM~2.5~ value for each paper. But, there are papers where more than one mean PM~2.5~ value is reported (for example, one for the male group and one for the female group). In these cases, we have picked one value from the ones that are available.


  * In some papers, the exact start and/or end year of a study is unclear. In these cases, we have mostly chosen the first instance of the multiple “study duration ranges” (depending on multiple plausible interpretations of when the final follow up ended) the paper.

  * In many papers, the upper limit or lower limit for age is not precisely specified. In these cases, we have recorded a NA. For example, there are papers where the upper limit category for age is 85+. In these cases, although we know the upper limit category, we still don't know the upper limit of the age, which could be 90, 95, 100, etc.
  
  * In places, where the cohort size is specifed in terms of "number of regions" (for example, 17 districts) and not "number of people", we have recorded a NA in the cohort size column.
  

  * Under the “methods” column (in the master dataset), we have only mentioned a subset of methods that were used to carry out different parts of the study. Authors may have used other methods than those mentioned in our meta analysis.
  
  * Different graphs are generated using different subsets of the [analysis dataset](https://docs.google.com/spreadsheets/d/1AljEJhNPLWX_8xRbT_HJuERBpbQt_QGixgJ9jEzFyQw/edit#gid=2082201996). Most of them are PM~2.5~ specific (our main focus), others include all pollutants (PM~2.5~, PM~10~, TSP, Ultrafine Particles). The type of papers used to generate a given graph is specified within the graph and/or the accompanying text.
  
  
## Results

&nbsp;


### PM~2.5~ exposure range and the Global Population Distribution

&nbsp;

```{r mean_pm2.5_population_graph_calcs, echo=FALSE, message=FALSE, warning=FALSE}

#> Add a note mentioning that this graph is PM2.5 specific (obvious from the graph)


# aqli color population in ordered pm2.5 buckets
aqli_color_grp_pm2.5_buckets <- aqli_color %>%
  mutate(region = ifelse(pm2020 >= 0 & pm2020 <= 25, "0-25", pm2020), 
         region = ifelse(pm2020 > 25 & pm2020 <= 50, ">25-50", region), 
         region = ifelse(pm2020 > 50 & pm2020 <= 75, ">50-75", region), 
         region = ifelse(pm2020 > 75 & pm2020 <= 100, ">75-100", region), 
         region = ifelse(pm2020 > 100, ">100", region)) %>%
  group_by(region) %>%
  summarise(tot_pop = sum(population, na.rm = TRUE)) %>%
  ungroup() %>%
    mutate(order_pollution_group = ifelse(region == "0-25", 1, 0), 
         order_pollution_group = ifelse(region == ">25-50", 2, order_pollution_group), 
         order_pollution_group = ifelse(region == ">50-75", 3, order_pollution_group), 
         order_pollution_group = ifelse(region == ">75-100", 4, order_pollution_group), 
         order_pollution_group = ifelse(region == ">100", 5, order_pollution_group)) %>%
  ungroup() %>%
  mutate(tot_pop_prop = (tot_pop/sum(tot_pop))*100)
  
  
# epi total number of studies in ordered pollution buckets (makes sense to count different countries in a pooled study as different studies)
epi_num_studies_grp_pm2.5_buckets <- epi %>%
  filter(!is.na(mean_pm2.5) , mean_pm2.5 != "NA", non_pm2.5 == 0) %>%
    mutate(region = ifelse(mean_pm2.5 >= 0 & mean_pm2.5 <= 25, "0-25", mean_pm2.5), 
         region = ifelse(mean_pm2.5 > 25 & mean_pm2.5 <= 50, ">25-50", region), 
         region = ifelse(mean_pm2.5 > 50 & mean_pm2.5 <= 75, ">50-75", region), 
         region = ifelse(mean_pm2.5 > 75 & mean_pm2.5 <= 100, ">75-100", region), 
         region = ifelse(mean_pm2.5 > 100, ">100", region)) %>%
  mutate(order_pollution_group = ifelse(region == "0-25", 1, 0), 
         order_pollution_group = ifelse(region == ">25-50", 2, order_pollution_group), 
         order_pollution_group = ifelse(region == ">50-75", 3, order_pollution_group), 
         order_pollution_group = ifelse(region == ">75-100", 4, order_pollution_group), 
         order_pollution_group = ifelse(region == ">100", 5, order_pollution_group)) %>%
  group_by(region) %>%
  summarise(tot_studies = n(), order_pollution_group = order_pollution_group[1]) %>%
  ungroup() %>%
  mutate(tot_studies_prop = (tot_studies/sum(tot_studies))*100)


#> line plot version (commented for now, can uncomment later if required)
# ggplot(mapping = aes(x= fct_reorder(region, order_pollution_group), group = 1)) +
# geom_smooth(data = aqli_color_grp_pm2.5_buckets, mapping = aes(y = tot_pop_prop, color = "Percent of World Population")) +
# geom_smooth(data = epi_num_studies_grp_pm2.5_buckets, mapping = aes(y = tot_studies_prop, color = "Percent of Total Studies")) +
#   labs(x = "pm2.5 buckets", y = "Percent of World Population") +
#   scale_y_continuous(sec.axis = sec_axis(~.*1, name = "Percent of Total Studies")) +
#   scale_color_manual(name = "", breaks = c("Percent of World Population", "Percent of Total Studies"), values = c("Percent of World Population" = "blue", "Percent of Total Studies" = "red")) +
# ggthemes::theme_hc()

#> bar chart version

# creating a master dataset for both population and studies data
pop_epi_studies_data <- aqli_color_grp_pm2.5_buckets %>%
  left_join(epi_num_studies_grp_pm2.5_buckets, by = c("region", "order_pollution_group")) %>%
  select(region,  order_pollution_group, tot_pop_prop, tot_studies_prop) %>%
  pivot_longer(cols = tot_pop_prop:tot_studies_prop, names_to = "type_of_prop", values_to = "val")

# plotting the bar graph
pop_num_studies_in_pollution_buckets_graph <- pop_epi_studies_data %>%
  ggplot() +
  geom_col(mapping = aes(x = fct_reorder(region, order_pollution_group), y = val, fill = type_of_prop), position = position_dodge(), width = 0.7) +
  scale_y_continuous(breaks = seq(0, 100, 10)) +
  scale_fill_manual(values = c("tot_pop_prop" = "grey", "tot_studies_prop" = "cornflowerblue"), labels = c("Proportion of World Population in Bucket", "Proportion of Studies Completed in Bucket")) + 
  labs(x = "Mean PM2.5 bucket (in µg/m³)",  y = "Percentage", fill = "") +
  theme_hc() +
  theme(axis.line.y = element_line(color = "black"), 
        axis.line.x = element_line(color = "black"))


#> mean pm2.5 distribution segregated by country (density plot): makes sense to count different country studies separately, as we are making country wise distributions, so we want to see all PM2.5 ranges covered in a given country.
mean_pm2.5_country_wise_graph <- epi %>% 
  filter(!is.na(mean_pm2.5), mean_pm2.5 != "NA", non_pm2.5 == 0) %>%
  ggplot() +
  geom_density(mapping = aes(x = mean_pm2.5, fill = country), alpha = 0.5, color = "black", position = "identity") +
  labs(x = expression(paste("Mean PM2.5 concentration (", mu, "g", "/", m^3, ")"))) +
  theme_tufte()


#> mean pm2.5 distribution segregated by country (histogram)
# epi %>%
#   filter(!is.na(mean_pm2.5) | mean_pm2.5 != "NA") %>%
#   ggplot() +
#   geom_histogram(mapping = aes(x = mean_pm2.5, fill = country), alpha = 0.5, color = "white", position = "identity") +
#   scale_x_continuous(breaks = seq(1, 100, 2)) +
#   labs(x = expression(paste("Mean PM2.5 concentration (", mu, "g", "/", m^3, ")"))) +
#   theme_tufte()

#> other calculations for this block of code

#> total pop where pollution is > 50 micrograms per cubic meter 

# thresholds
pm2.5_thresh_highly_polluted <- 50
pm2.5_thresh_severly_polluted <- 75
pm2.5_thresh_low_polluted <- 25

# higher than hp

# total pop where pollution is > x micrograms per cubic meter (hp)
total_pop_above_x_hp_micr_gm <- aqli_color %>%
  filter(pm2020 > pm2.5_thresh_highly_polluted) %>%
  summarise(tot_pop = sum(population, na.rm = TRUE)) %>%
  unlist()

# percent pop where pollution is > x micrograms per cubic meter (hp)
perc_pop_above_x_hp_micr_gm <- round((total_pop_above_x_hp_micr_gm/world_pop)*100, 1)
  
# total pop where pollution is > x micrograms per cubic meter in millions (hp)
tot_pop_above_x_hp_micr_gm_millions <- round(total_pop_above_x_hp_micr_gm/1000000, 1)

# higher than sp

# total pop where pollution is > x micrograms per cubic meter (sp)
total_pop_above_x_sp_micr_gm <- aqli_color %>%
  filter(pm2020 > pm2.5_thresh_severly_polluted) %>%
  summarise(tot_pop = sum(population, na.rm = TRUE)) %>%
  unlist()

# percent pop where pollution is > x micrograms per cubic meter (sp)
perc_pop_above_x_sp_micr_gm <- round((total_pop_above_x_sp_micr_gm/world_pop)*100, 1)
  
# total pop where pollution is > x micrograms per cubic meter in millions (sp)
tot_pop_above_x_sp_micr_gm_millions <- round(total_pop_above_x_sp_micr_gm/1000000, 1)

# less than lp

# total pop where pollution is < x micrograms per cubic meter (lp)
total_pop_below_x_lp_micr_gm <- aqli_color %>%
  filter(pm2020 < pm2.5_thresh_low_polluted) %>%
  summarise(tot_pop = sum(population, na.rm = TRUE)) %>%
  unlist()

# percent pop where pollution is < x micrograms per cubic meter (lp)
perc_pop_below_x_lp_micr_gm <- round((total_pop_below_x_lp_micr_gm/world_pop)*100, 1)
  
# total pop where pollution is > x micrograms per cubic meter in millions (lp)
tot_pop_below_x_lp_micr_gm_millions <- round(total_pop_below_x_lp_micr_gm/1000000, 1)


#> percent studies in a given pollution bucket 


# higher than hp: makes sense to count different countries in a pooled study as different studies.
epi_num_studies_in_hp <- epi %>%
  filter(!is.na(mean_pm2.5) , mean_pm2.5 != "NA", non_pm2.5 == 0) %>%
  filter(mean_pm2.5 > pm2.5_thresh_highly_polluted) %>%
  nrow() %>%
  unlist()

perc_epi_studies_in_hp <- round((epi_num_studies_in_hp/nrow(epi %>% filter((!is.na(mean_pm2.5) | mean_pm2.5 != "NA"), non_pm2.5 == 0)))*100, 1)

# higher than sp: makes sense to count different countries in a pooled study as different studies.

epi_num_studies_in_sp <- epi %>%
  filter(!is.na(mean_pm2.5), mean_pm2.5 != "NA", non_pm2.5 == 0) %>%
  filter(mean_pm2.5 > pm2.5_thresh_severly_polluted) %>%
  nrow() %>%
  unlist()

perc_epi_studies_in_sp <- round((epi_num_studies_in_sp/nrow(epi %>% filter(!is.na(mean_pm2.5), mean_pm2.5 != "NA", non_pm2.5 == 0)))*100, 1)

# less than lp: makes sense to count different countries in a pooled study as different studies.

epi_num_studies_in_lp <- epi %>%
  filter(!is.na(mean_pm2.5), mean_pm2.5 != "NA", non_pm2.5 == 0) %>%
  filter(mean_pm2.5 < pm2.5_thresh_low_polluted) %>%
  nrow() %>%
  unlist()

perc_epi_studies_in_lp <- round((epi_num_studies_in_lp/nrow(epi %>% filter(!is.na(mean_pm2.5), mean_pm2.5 != "NA", non_pm2.5 == 0)))*100, 1)

# plot population and number of studies bar graph
pop_num_studies_in_pollution_buckets_graph

# plot mean pm2.5 country wise graph
mean_pm2.5_country_wise_graph
```

&nbsp;

  * In total `r length(unique(epi$paper_uid))` AQI epi studies were included in the final [analysis dataset](https://docs.google.com/spreadsheets/d/1AljEJhNPLWX_8xRbT_HJuERBpbQt_QGixgJ9jEzFyQw/edit#gid=2082201996).
  
  
  * `r perc_pop_above_x_hp_micr_gm` percent of the world population, or `r tot_pop_above_x_hp_micr_gm_millions` million people, live in areas where the annual average PM~2.5~ pollution is greater than `r pm2.5_thresh_highly_polluted` µg/m³. But, only `r perc_epi_studies_in_hp` percent (`r epi_num_studies_in_hp` PM~2.5~ specific studies) of the total studies, have been performed in these highly polluted parts of the world. These highly polluted areas are areas where the average PM~2.5~ pollution is at least `r round(pm2.5_thresh_highly_polluted/who_pm2.5_guideline, 1)` times the WHO PM~2.5~ safe guideline of `r who_pm2.5_guideline` µg/m³.


  * Approximately `r perc_pop_above_x_sp_micr_gm` percent of the world population (`r tot_pop_above_x_sp_micr_gm_millions` million people) live in the most severely polluted parts of the world, where annual average PM~2.5~ pollution is upwards of `r pm2.5_thresh_severly_polluted` µg/m³ (at least `r round(pm2.5_thresh_severly_polluted/who_pm2.5_guideline, 1)` times the WHO safe guideline). In these most severely polluted parts of the world, `r epi_num_studies_in_sp` AQ epi studies have been performed .


  * Most of the AQ epi studies (`r perc_epi_studies_in_lp` percent of the total, or `r epi_num_studies_in_lp` studies) performed so far, are concentrated in areas where the average PM~2.5~ concentration is in the 0-25 µg/m³ range. People living in these areas (`r perc_pop_below_x_lp_micr_gm` percent of the world population) are breathing air that is much less polluted relative to the people living in the most polluted parts of the world (as seen above). But, even in the 0-25 µg/m³ bucket, anyone living above `r who_pm2.5_guideline` µg/m³, is out of compliance with the WHO PM~2.5~ guideline.
  
  * Note: In making the above 2 graphs, multiple countries of a pooled study are counted as separate studies.
  
&nbsp;

### Geographic Distribution of Studies


```{r geographic_dist_studies_calcs, echo=FALSE, message=FALSE, warning=FALSE}

#> Add a note mentioning that the underlying data for this graph includes all types of pollutants and also that, each country in a pooled study is counted separately.

# generate summary dataset for total number of studies conducted in a given country: (makes sense to count each country in a pooled dataset)
epi_country_count <- epi %>%
  select(country) %>%
  count(country) %>%
  filter(country != "NA", !is.na(country), n != "NA", !is.na(n)) 

# making country names match before joining epi file with color file
epi_country_count$country <- str_replace(epi_country_count$country, "USA", "United States")

#> plot choropleth map

# get data ready for the choropleth map
world <- ne_countries(scale = "medium", returnclass = "sf")


# making country names match before joining epi file with color file
world$name_long <- str_replace(world$name_long, "Russian Federation", "Russia")
world$name_long <- str_replace(world$name_long, "Lao PDR", "Laos")
world$name_long <- str_replace(world$name_long, "Dem. Rep. Korea", "North Korea")
world$name_long <- str_replace(world$name_long, "Republic of Korea", "South Korea")
world$name_long <- str_replace(world$name_long, "Vatican", "Vatican City")
world$name_long <- str_replace(world$name_long, "Brunei Darussalam", "Brunei")
world$name_long <- str_replace(world$name_long, "Somaliland", "Somalia")

# joining world shape file with the epi country wise count dataset
world_shp_epi <- world %>%
  left_join(epi_country_count, by = c("name_long" = "country")) %>%
  select(name_long, n, geometry) %>%
  rename(num_studies = n, country = name_long)

# color 2020 collapse to country level to get the PM2.5 pollution layer (first layer of the map)
aqli_color_country <- aqli_color %>%
  group_by(country) %>%
  mutate(pop_weights = population/sum(population, na.rm = TRUE), 
         pm2020_pop_weighted = pm2020*pop_weights) %>%
  summarise(avg_pm2.5_2020 = sum(pm2020_pop_weighted, na.rm = TRUE))

# joining the world shape + epi joined file with the aqli color file 
world_shp_epi_color <- world_shp_epi %>%
  left_join(aqli_color_country, by = "country") %>%
  filter(country != "Antarctica") %>%
  select(country, num_studies, avg_pm2.5_2020, geometry)

# In the total number of studies column, replacing NAs with 0s
world_shp_epi_color <- world_shp_epi_color %>%
  mutate(num_studies = ifelse(is.na(num_studies) == TRUE, 0, num_studies))

# getting centroids for countries after correcting for problematic polygons
world_shp_epi_color_for_centroids <- st_make_valid(world_shp_epi_color)
country_wise_centroids <- st_centroid(world_shp_epi_color_for_centroids, of_largest_polygon = TRUE)

# plotting the choropleth
geographic_dist_graph <- world_shp_epi_color %>%
  ggplot() +
  geom_sf(mapping = aes(fill = avg_pm2.5_2020), color = "white") +
 scale_fill_continuous_sequential(palette = "YlOrRd") +
  geom_sf(data = country_wise_centroids %>% filter(num_studies != 0), mapping = aes(size = num_studies), col = "grey", ) +
    theme(legend.position = "bottom") +
  scale_size_binned() +
  scale_size(range = c(0, 8)) +
  theme_map() +
  theme(legend.position = "bottom")

#> Other calculations for this block of code

# total number of studies performed in USA and Canada
num_studies_usa_canada <- epi_country_count %>% 
  filter(country %in% c("United States", "Canada")) %>%
  summarise(total_studies = sum(n, na.rm = TRUE)) %>%
  unlist()

# total number of studies performed in USA and Canada and Europe
num_studies_usa_canada_europe <- epi %>% 
  filter((country %in% c("Canada", "USA")) | (continent %in% c("Europe"))) %>% 
  nrow()

# areas other than USA, Canada and Europe
areas_other_than_usa_canada_europe <- epi %>% 
  filter((country %notin% c("Canada", "USA")) & (continent %notin% c("Europe"))) %>%
  select(country) %>%
  unique() %>%
  unlist() %>%
  as.vector()

# continent order list (order variable: number of studies)
continent_wise_study_summary <- epi %>% 
  group_by(continent) %>% 
  summarise(count = n(), average_study_duration = mean(study_duration, na.rm = TRUE)) %>%
  slice_max(count, n = 6) %>%
  mutate(prop_studies = round((count/sum(count, na.rm = TRUE))*100, 1))

# total number of studies in South America
num_studies_south_america <- epi %>%
  filter(continent == "South America") %>%
  nrow()

# total number of studies in Oceania
num_studies_oceania <- epi %>%
  filter(continent == "Oceania") %>%
  nrow()

# total number of studies in Africa
num_studies_africa <- epi %>%
  filter(continent == "Africa") %>%
  nrow()

# plot geographic distribution
geographic_dist_graph

```

  * `r num_studies_usa_canada` (`r round((num_studies_usa_canada/nrow(epi))*100, 1)` percent) studies were performed in the USA and Canada. 
  
  
  * `r num_studies_usa_canada_europe` studies (`r round((num_studies_usa_canada_europe/nrow(epi))*100, 1)` percent) were performed in the USA, Canada and Europe combined. Remaining `r 100 - round((num_studies_usa_canada_europe/nrow(epi))*100, 1)` percent studies are scattered across `r areas_other_than_usa_canada_europe`.
  
  
  * In terms of the total number of studies conducted, `r continent_wise_study_summary$continent[1]` dominates the rest of the continents (with `r continent_wise_study_summary$prop_studies[1]` percent of total studies). This is closely followed by `r continent_wise_study_summary$continent[2]` (with `r continent_wise_study_summary$prop_studies[2]` percent of total studies) and `r continent_wise_study_summary$continent[3]`  (with `r continent_wise_study_summary$prop_studies[3]` percent of total studies). `r continent_wise_study_summary$continent[4:6]` combined, have only seen `r num_studies_south_america + num_studies_oceania + num_studies_africa` studies.
  
  * Note: In making the above graph, multiple countries of a pooled study are counted as separate studies. Also, the above graph represents all studies (i.e. includes all pollutants and is not just PM~2.5~ specific).
  
  * **Note for discussion: In the total count of studies, each country in a multi-country is counted as separate studies.**

&nbsp;
  
### AQ Epi studies over time

&nbsp;

```{r aq_epi_studies_over_time_graph, echo=FALSE, message=FALSE, warning=FALSE}

#> Add a note mentioning that the underlying data for this graph includes all types of pollutants but only one row is counted for a given pooled (multi-country) study.

epi %>% 
dplyr::distinct(paper_uid, .keep_all = TRUE) %>%
  group_by(publishing_year) %>%
  summarise(total_studies = n()) %>%
  ungroup() %>%
  filter(total_studies != "NA", !is.na(total_studies), publishing_year != "NA", !is.na(publishing_year)) %>%
  ggplot() +
  geom_line(mapping = aes(x = publishing_year, y = total_studies), lwd = 1.2, color = "cornflowerblue") +
  labs(x = "Year", y = "Total Number of Studies") +
  theme_minimal()

```



  * In most of the 90’s and early 2000’s, the rate of AQ epi studies publishing was around 1 to 2 studies per annum on average.

  * Post 2009, there has been a noticeable increase in the overall volume of AQ epi studies published.
  
  * **Note for discussion: In making the above graph, multiple countries of a pooled study are *not* counted separately, i.e. a pooled multi-country study is counted as a single study. Also, the above graph represents all studies (i.e. includes all pollutants and is not just PM~2.5~ specific).**



### Distribution of duration of study by continent

&nbsp;

```{r dist_duration_study_calcs, echo=FALSE, message=FALSE, warning=FALSE}

#>  In this, in the graph multiple countries of a pooled study are counted separately. Also, this is not a PM2.5 specific graph.

#> Density plot of study duration (commented for now, can use later if required)
# epi %>%
#   filter(!is.na(study_duration)) %>%
#   ggplot() +
#   geom_density(mapping = aes(x = study_duration, fill = continent), alpha = 0.5) +
#     labs(x = "Study Duration") +
#   ggtitle("Epi Studies Duration Distribution (Continent Wise): Density Plot") +
#   scale_x_continuous(breaks = seq(0, 40, 5)) +
#   theme_tufte() +
#   theme(legend.position = "bottom")


#> Histogram of study duration

# adding an order panel 
continent_wise_study_duration_graph <- epi %>%
  dplyr::filter(continent != "NA", !is.na(continent)) %>%
  mutate(order_continent = ifelse(continent == "Asia", 1, 0), 
         order_continent = ifelse(continent == "Europe", 2, order_continent), 
         order_continent = ifelse(continent == "North America", 3, order_continent), 
         order_continent = ifelse(continent == "South America", 4, order_continent), 
         order_continent = ifelse(continent == "Africa", 5, order_continent),
         order_continent = ifelse(continent == "Oceania", 6, order_continent)) %>% 
  ggplot() +
  geom_histogram(mapping = aes(x = study_duration, fill = continent), alpha = 0.5, color = "white", position = "identity", binwidth = 2) +
    labs(x = "Study Duration") +
  ggtitle("Epi Studies Duration Distribution (Continent Wise): Histogram") +
  scale_x_continuous(breaks = seq(0, 40, 10)) +
  theme_hc() + 
  theme(legend.title = element_blank()) +
   facet_wrap(~fct_reorder(continent, order_continent), nrow = 1) +
  theme(axis.line.y = element_line(color = "black"), 
        axis.line.x = element_line(color = "black")) 

#> calculations relevant for this block of code

# long term study threshold
long_term_study_duration <- 25

# countries with studies greater than x (long_term) years of study duration
epi_long_term_studies_countries <- epi %>%
  filter(study_duration > long_term_study_duration, study_duration != "NA", !is.na(study_duration)) %>%
  select(country) %>%
  unlist() %>%
  unique() 


# continent order list (order variable: number of studies): code present in the geographical distribution map block.

# plot continent wise study duration graph
continent_wise_study_duration_graph


```

  * All of the “really” long term studies (> `r long_term_study_duration` years) ever conducted are concentrated only in `r epi_long_term_studies_countries`.

  * `r continent_wise_study_summary$continent[1:3]` have seen a relatively large number of studies (across a wide range of study durations). In contrast to this, in `r continent_wise_study_summary$continent[4:6]` combined, only `r num_studies_south_america + num_studies_oceania + num_studies_africa` studies have been performed.  


 * **Note for discussion: In making the above graph, multiple countries of a pooled study are counted separately. Also, the above graph represents all studies (i.e. includes all pollutants and is not just PM~2.5~ specific).**

### Major takeaways

  * `r round((num_studies_usa_canada/nrow(epi))*100, 1)` percent of the total number of studies are concentrated in the USA and Canada alone. 


  * Aproximately `r perc_pop_above_x_sp_micr_gm` percent of the world population, or `r tot_pop_above_x_sp_micr_gm_millions` million people live in areas where PM~2.5~ pollution concentrations are upwards of `r pm2.5_thresh_severly_polluted` µg/m³ (at least `r pm2.5_thresh_severly_polluted/who_pm2.5_guideline` times the WHO safe guideline). In these parts of the world, `r epi_num_studies_in_sp` AQ epi studies have been performed.


  * All of the “really” long term studies (> `r long_term_study_duration` years) ever conducted are concentrated only in `r epi_long_term_studies_countries`.


  * `r continent_wise_study_summary$continent[4:6]` combined, have only seen `r num_studies_south_america + num_studies_oceania + num_studies_africa` studies.


  * In the last decade, there has been a noticeable increase in the total number of AQ epi studies performed. But, most of these new studies are concentrated in the low pollution concentration ranges of 0-25 µg/m³. The high pollution concentration ranges remain poorly explored.


### Conclusion

All studies included in this analysis point to the same overall picture:air pollution is a serious health threat. The existing state of scientific literature on air pollution and health is clear that air pollution’s impact on health is well-established and taking action in a polluted environment should not be delayed in order to complete multi-year large cohort epidemiological studies in an area, even if there has not been a prior study in that particular geography. That said, it is important for the field of air quality epidemiolgy to understand the contours of its current research landscape to most effectively identify directions for future research and deploy limited resources. 

&nbsp;


### Methodology 

Through a comprehensive and ongoing* literature review, we are making an attempt at creating an exhaustive public listing of all the epidemiological studies out there (that we could find) that examine the relationship between PM~2.5~ and Life Expectancy/Mortality. 

For each study, we record data on key defining features, such as: Geography, Cohort Size, Study Duration, PM~2.5~ exposure range, etc. Then we used [this](https://docs.google.com/spreadsheets/d/1AljEJhNPLWX_8xRbT_HJuERBpbQt_QGixgJ9jEzFyQw/edit#gid=2082201996) analysis dataset to carry out a meta-analysis, results of which are detailed in the Results section above.

We are seeking to make this analysis as current, complete and error-free as possible and view it as a continual work in progress. We would welcome the air quality community’s any comments, corrections, andor suggestions. Please contact **aqli-epic@uchicago.edu** or leave a comment in this GitHub repository. 

### Limitations (To be discussed)


&nbsp;


### How can you (the community) help in improving this analysis?


  * Add to the [**__analysis dataset__**](https://docs.google.com/spreadsheets/d/1AljEJhNPLWX_8xRbT_HJuERBpbQt_QGixgJ9jEzFyQw/edit#gid=2082201996) and and grow the epi database:


  * If you know of other papers that: (a) are attempting to study the link between PM~2.5~ and Life Expectancy/Mortality and (b) are not included in this analysis: Please leave a comment in the [**__analysis dataset__**](https://docs.google.com/spreadsheets/d/1AljEJhNPLWX_8xRbT_HJuERBpbQt_QGixgJ9jEzFyQw/edit#gid=2082201996), providing a link to the paper. You can also write to us at aqli-info@uchicago.edu (with the link to the paper mentioned in the email). 


    * As a next step, we’ll go through your submission, and if it fits our inclusion criteria, we will update the underlying analysis dataset and re-render the entire blog, so that it represents the most up to date data and figures.
    
    * If you are unsure (given the inclusion criteria) about whether a particular paper (that you want to post) should be posted on not, we encourage you to post it and let us worry about the inclusion/exclusion bit. 
    

  * In case you have comments on some aspects of a paper/if you find any errors, please leave a comment in the [**__analysis dataset__**](https://docs.google.com/spreadsheets/d/1AljEJhNPLWX_8xRbT_HJuERBpbQt_QGixgJ9jEzFyQw/edit#gid=2082201996) on the cell (tagging aqli-info@uchicago.edu, using **@** symbol) where the error is found or write to us at aqli-info@uchicago.edu, detailing the error and its location (cell address on the sheet).
    

&nbsp;

&nbsp; 

<hr>

### References



&nbsp;

&nbsp;

<hr>

### Other Graphs and Interactive Summary Dashboard

To further explore these graphs and more in an interactive setup, visit the [AQ Epi dashboard](https://aarsh.shinyapps.io/aqli-epic-epi-meta-analysis/).


```{r pm2.5_expo_ul_ll, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}

####  PM2.5 distributions of the lower limits and upper limits of exposure range (Density and Histogram Plots)

## All graphs below this block of code is deactivated for the blog post. This will be displayed on the dashboard, alongside other graphs.


#> Specific to PM2.5 and count separate countries in a pooled study separately

#> exposure distribution density and histogram plots (PM2.5 LL and PM2.5 UL)

# create long dataset
epi_long <- epi %>% 
  filter(!is.na(mean_pm2.5), mean_pm2.5 != "NA", non_pm2.5 == 0) %>%
  pivot_longer(cols = contains("pm2.5_exposure"), names_to =  "exposure_type", values_to = "exposure_value") %>%
  select(exposure_type, exposure_value) %>%
  filter(!is.na(exposure_value)) 

# density plot
epi_long %>%
  ggplot() +
  geom_density(mapping = aes(x = exposure_value, fill = exposure_type), alpha = 0.6, position = "identity") +
  labs(x = expression(paste("PM2.5 concentration (", mu, "g", "/", m^3, ")"))) +
  theme_tufte()

# histogram plot
epi_long %>%
  ggplot() +
  geom_histogram(mapping = aes(x = exposure_value, fill = exposure_type), position = "identity", alpha = 0.4, color = "white") +
  labs(x = expression(paste("PM2.5 concentration (", mu, "g", "/", m^3, ")"))) +
  theme_tufte()


```



```{r age_dist_ul_ll, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}

#### Age distributions of lower limits and upper limits of age range (Density and Histogram plots)

#> Different countries in a pooled study should be counted separately. As of now this is PM2.5 specific, but can be generalized in the future.

#> (5) age distribution plot (LL and UL) using the long dataset (density plot)
epi %>% 
  filter(!is.na(mean_pm2.5), mean_pm2.5 != "NA", non_pm2.5 == 0, (!is.na(cohort_age_ll) | !is.na(cohort_age_ul) | cohort_age_ll != "NA" | cohort_age_ul != "NA")) %>%
  pivot_longer(cols = contains("cohort_age"), names_to =  "age_dist_type", values_to = "age_value") %>%
  select(age_dist_type, age_value) %>%
  filter(!is.na(age_value)) %>%
  ggplot() +
  geom_density(mapping = aes(x = age_value, fill = age_dist_type), alpha = 0.6, color = "black") +
  labs(x = "Age") +
  theme_tufte()

#> (6) age distribution plot (LL and UL) using the long dataset (histogram plot)
epi %>% 
  filter(!is.na(mean_pm2.5), mean_pm2.5 != "NA", non_pm2.5 == 0, (!is.na(cohort_age_ll) | !is.na(cohort_age_ul) | cohort_age_ll != "NA" | cohort_age_ul != "NA")) %>%
  pivot_longer(cols = contains("cohort_age"), names_to =  "age_dist_type", values_to = "age_value") %>%
  select(age_dist_type, age_value) %>%
  filter(!is.na(age_value)) %>%
  ggplot() +
  geom_histogram(mapping = aes(x = age_value, fill = age_dist_type), alpha = 0.5, position = "identity", color = "white") +
  labs(x = "Age") 

```


```{r country_wise_cohort_size, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}

#### Country wise distribution of Cohort Sizes (Density and Histogram plots)

#> Different countries in a pooled study should be counted separately. As of now this is PM2.5 specific, but can be generalized in the future.


#> (7) Country Wise Cohort Size density plot
epi %>%
    filter(!is.na(mean_pm2.5), mean_pm2.5 != "NA", non_pm2.5 == 0, !is.na(cohort_size), cohort_size != "NA") %>%
  ggplot() +
  geom_density(mapping = aes(x = log10(cohort_size), fill = country), alpha = 0.5) +
  theme_tufte()

#> (8) Country wise Cohort Size histogram plot
epi %>%
  filter(!is.na(mean_pm2.5), mean_pm2.5 != "NA", non_pm2.5 == 0, !is.na(cohort_size), cohort_size != "NA") %>%
  ggplot() +
  geom_histogram(mapping = aes(x = log10(cohort_size), fill = country), alpha = 0.5, position = "identity") +
  theme_tufte()

```


```{r study_duration_dist_country_wise, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}

#### Country wise Distribution of Study Duration (Density and Histogram Plots)

#> Different countries in a pooled study should be counted separately. As of now this is PM2.5 specific, but can be generalized in the future.


#> (9) Country wise Study duration density plot
epi %>%
  filter(!is.na(mean_pm2.5), mean_pm2.5 != "NA", non_pm2.5 == 0, !is.na(study_duration), study_duration != "NA") %>%
  ggplot() +
  geom_density(mapping = aes(x = study_duration, fill = country, alpha = 0.5)) +
  theme_tufte()

#> Country Wise Study Duration histogram

epi %>%
  filter(!is.na(mean_pm2.5), mean_pm2.5 != "NA", non_pm2.5 == 0, !is.na(study_duration), study_duration != "NA") %>%
  ggplot() +
  geom_histogram(mapping = aes(x = study_duration, fill = country, alpha = 0.5), position = "identity", alpha = 0.5, color = "white", binwidth = 1) +
  theme_tufte() +
  facet_wrap(~country, nrow = 5) +
  theme(legend.position = "bottom")


  




```




