---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Epidemiological Studies: Meta Analysis

<!-- badges: start -->
<!-- badges: end -->

The goal of epi.meta.analysis is to:

* Exhaustively Explore epidemiological research studies to figure out the regional (country wise) PM2.5 exposure range covered in these studies.
* Identify and plot the locations of these studies on a map to identify locations where no/very few cohort studies have been undertaken.
* Better understand the relationship between PM2.5 and Mortality/Life Expectancy.



```{r load_libraries, message=FALSE, warning=FALSE, include=FALSE}
library(readr)
library(dplyr)
library(stringr)
library(magrittr)
library(ggplot2)
library(readr)
library(tidytext)
library(tidyr)
library(tidyverse)
library(ggthemes)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
```


# Analysis and Plots 

* Note: All data that is used to generate the graphs below can be found in the `data-raw` sub-directory, which is present at the root of the `epi.meta.analysis` directory.

```{r prepare_data, message=FALSE, warning=FALSE, include=FALSE}
#> read in datasets 

# epi studies analysis raw sheet
epi <- readxl::read_xlsx("./data-raw/pm2.5_distribution/AQLI_Epidemiology Literature Research.xlsx", sheet = "PM2.5MortalityAnalysisDataset")

# AQLI color file
aqli_color <- read_csv("./data-raw/pm2.5_distribution/color.csv") 


#> change default columns types
epi$cohort_size <- as.numeric(epi$cohort_size)
epi$study_start_year<- as.numeric(epi$study_start_year)
epi$study_end_year <- as.numeric(epi$study_end_year)
epi$pm2.5_exposure_ll <- as.numeric(epi$pm2.5_exposure_ll)
epi$pm2.5_exposure_ul <- as.numeric(epi$pm2.5_exposure_ul)
epi$mean_pm2.5 <- as.numeric(epi$mean_pm2.5)
epi$sd_pm2.5 <- as.numeric(epi$sd_pm2.5)
epi$cohort_age_ll <- as.numeric(epi$cohort_age_ll)
epi$cohort_age_ul <- as.numeric(epi$cohort_age_ul)

#> add useful columns

epi <- epi %>%
  mutate(study_duration = (study_end_year - study_start_year) + 1)
```


##  PM2.5 distributions of the lower limits and upper limits of exposure range (Density and Histogram Plots)

```{r pm2.5_expo_ul_ll, echo=FALSE, message=FALSE, warning=FALSE}
#> exposure distribution density and histogram plots (PM2.5 LL and PM2.5 UL)

# create long dataset
epi_long <- epi %>% 
  pivot_longer(cols = contains("pm2.5_exposure"), names_to =  "exposure_type", values_to = "exposure_value") %>%
  select(exposure_type, exposure_value) %>%
  filter(!is.na(exposure_value)) 

# density plot
epi_long %>%
  ggplot() +
  geom_density(mapping = aes(x = exposure_value, fill = exposure_type), alpha = 0.6, position = "identity") +
  scale_x_continuous(breaks = seq(0, 250, 10)) + 
  labs(x = expression(paste("PM2.5 concentration (", mu, "g", "/", m^3, ")"))) +
  theme_tufte()

# histogram plot
epi_long %>%
  ggplot() +
  geom_histogram(mapping = aes(x = exposure_value, fill = exposure_type), position = "identity", alpha = 0.4) +
  scale_x_continuous(breaks = seq(0, 250, 10)) + 
  labs(x = expression(paste("PM2.5 concentration (", mu, "g", "/", m^3, ")"))) +
  theme_tufte()



```

* From **Plot 1**, it looks like most of the studies are present in concentration ranges that are less than 50 µg/m³. According to AQLIs latest estimates, 12.6 percent (~962 million people) of the world population live in areas where PM2.5 pollution concentration is greater than 50 µg/m³. 


## Country Wise Mean PM2.5 distribution (Density Plot and Historgram)
```{r mean_pm2.5_country_wise, echo=FALSE, message=FALSE, warning=FALSE}
#> mean PM2.5 distribution segregated by country

# (2) mean pm2.5 distribution segregated by country (density plot)
epi %>% 
  filter(!is.na(mean_pm2.5) | mean_pm2.5 != "NA") %>%
  ggplot() +
  geom_density(mapping = aes(x = mean_pm2.5, fill = country), alpha = 0.5, color = "black", position = "identity") +
  scale_x_continuous(breaks = seq(1, 100, 2)) +
  labs(x = expression(paste("Mean PM2.5 concentration (", mu, "g", "/", m^3, ")"))) +
  theme_tufte()

# (3) mean pm2.5 distribution segregated by country (histogram)
epi %>% 
  filter(!is.na(mean_pm2.5) | mean_pm2.5 != "NA") %>%
  ggplot() +
  geom_histogram(mapping = aes(x = mean_pm2.5, fill = country), alpha = 0.5, color = "white", position = "identity") +
  scale_x_continuous(breaks = seq(1, 100, 2)) + 
  labs(x = expression(paste("Mean PM2.5 concentration (", mu, "g", "/", m^3, ")"))) +
  theme_tufte()
  

```

## Choropleth world map: mapping the total number of epidemiological studies (specifically those that are trying to better understand the relationship between PM2.5 and Mortality/Life Expectancy).
```{r choropleth_study_count, echo=FALSE, message=FALSE, warning=FALSE}

#> (4) Choropleth world map mapping the count of total number of studies

# generate summary dataset for total number of studies conducted in a given country
epi_country_count <- epi %>%
  select(country) %>%
  count(country) %>%
  filter(country != "NA") 

epi_country_count$country <- str_replace(epi_country_count$country, "USA", "United States")


# plot choropleth map
world <- ne_countries(scale = "medium", returnclass = "sf")
world %>%
  left_join(epi_country_count, by = c("name_long" = "country")) %>%
ggplot() + 
  geom_sf(mapping = aes(fill = n)) +
  scale_fill_distiller(name = "Number of Studies",  palette = "OrRd", trans = "reverse") +
  theme_tufte()

```

## Age distributions of lower limits and upper limits of age range (Density and Histogram plots)
```{r plot4_age_distribution_ll_ul, echo=FALSE, message=FALSE, warning=FALSE}
#> (5) age distribution plot (LL and UL) using the long dataset (density plot)
epi %>% 
  pivot_longer(cols = contains("cohort_age"), names_to =  "age_dist_type", values_to = "age_value") %>%
  select(age_dist_type, age_value) %>%
  filter(!is.na(age_value)) %>%
  ggplot() +
  geom_density(mapping = aes(x = age_value, fill = age_dist_type), alpha = 0.6, color = "black") +
  scale_x_continuous(breaks = seq(0, 90, 5)) + 
  labs(x = "Age") +
  theme_tufte()

#> (6) age distribution plot (LL and UL) using the long dataset (histogram plot)
epi %>% 
  pivot_longer(cols = contains("cohort_age"), names_to =  "age_dist_type", values_to = "age_value") %>%
  select(age_dist_type, age_value) %>%
  filter(!is.na(age_value)) %>%
  ggplot() +
  geom_histogram(mapping = aes(x = age_value, fill = age_dist_type), alpha = 0.5, position = "identity", color = "white") +
  scale_x_continuous(breaks = seq(0, 90, 5)) + 
  scale_y_continuous(breaks = seq(0, 15, 2)) +
  labs(x = "Age") +
  theme_tufte()

```

## Country wise distribution of Cohort Sizes (Density and Histogram plots)
```{r plot5_cohort_size_dist_country_wise, echo=FALSE, message=FALSE, warning=FALSE}

#> (7) Country Wise Cohort Size density plot
epi %>%
  ggplot() +
  geom_density(mapping = aes(x = log10(cohort_size), fill = country), alpha = 0.5) +
  theme_tufte()

#> (8) Country wise Cohort Size histogram plot
epi %>%
  ggplot() +
  geom_histogram(mapping = aes(x = log10(cohort_size), fill = country), alpha = 0.5, position = "identity") +
  theme_tufte()

```


## Country wise Distribution of Study Duration (Density and Histogram Plots)
```{r study_duration_dist_country_wise, echo=FALSE, message=FALSE, warning=FALSE}

#> (9) Country wise Study duration density plot
epi %>%
  ggplot() +
  geom_density(mapping = aes(x = study_duration, fill = country, alpha = 0.5)) +
  scale_x_continuous(breaks = seq(0, 40, 5)) +
  theme_tufte()

#> (10) Country Wise Study Duration histogram

epi %>%
  ggplot() +
  geom_histogram(mapping = aes(x = study_duration, fill = country, alpha = 0.5), position = "identity", alpha = 0.5) +
  scale_x_continuous(breaks = seq(0, 40, 5)) +
  theme_tufte()

```


## AQLI data, top 10 most polluted countries: PM2.5 distributions Histograms and Summary Table
```{r pm2.5_country_dist_aqli_plot, echo=FALSE, message=FALSE, warning=FALSE}
# AQLI top 20 most polluted countries
n <- 10

top_10_countries_most_polluted <- aqli_color %>%
  group_by(country) %>%
  mutate(pop_weights = population/sum(population, na.rm = TRUE), 
         pm2020_pop_weighted = pm2020*pop_weights) %>%
  summarise(avg_pm2.5_2020 = sum(pm2020_pop_weighted)) %>%
  slice_max(avg_pm2.5_2020, n = n) 

top_10_countries_most_polluted_names_vec <- top_10_countries_most_polluted %>%
  select(country) %>%
  unlist() %>%
  as.vector()


#> (11) Top 10 most polluted countries: PM2.5 distributions
aqli_color %>%
  filter(country %in% top_10_countries_most_polluted_names_vec) %>%
  ggplot() +
  geom_histogram(mapping = aes(x = pm2020), color = "white") + 
  scale_x_continuous(breaks = seq(0, 120, 20)) + 
  facet_wrap(~country) +
  labs(x = expression(paste("PM2.5 concentration (", mu, "g", "/", m^3, ")"))) +
  theme_tufte()
```



### Summary Table

```{r pm2.5_country_dist_aqli_table, echo=FALSE, message=FALSE, warning=FALSE}
# Top 10 most polluted countries: PM2.5 distributions: Table
  
colnames(top_10_countries_most_polluted) <- c("Country" ,"Average PM2.5 2020 (µg/m³)")  
  
knitr::kable(top_10_countries_most_polluted)  









```

